Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Nearly finished writing up formal semantics.

\ADComment{TODO: write it up properly.}

\ADComment{TODO: remark that we do not model transmit faithfully.}

Thread state: $(l, \mathit{ss})$, where $l \in L$ is the thread's
local state, and $\mathit{ss}$ is a semi-colon-separated sequence of
program statements that the thread remains to execute.

Workgroup state: $((l_1, \mathit{ss}_1), \dots, (l_d,
\mathit{ss}_d))$, where each $(l_i, \mathit{ss}_i)$ is a thread state.

Kernel state: a pair $(\sigma, (w_1, w_2, \dots, w_M, \bot, \dots, \bot))$, where $\sigma$ is the shared state,
each $w_i$ is a workgroup state, which are
followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.

We treat thread-level semantics abstractly by assuming a transition
relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
\mathit{ss}'))$ that is defined unless $\mathit{ss} =
\mathsf{special}(); \mathit{tt}$, where $\mathsf{special}$ is one of
$\offerkill$, $\offerfork$, $\globalbarrier$ or
$\resizingglobalbarrier$.

To abstractly accounts for the interaction between global barrier
synchronization and the GPU memory model, we assume a function
$\mathsf{sync}$ that is applied to a global state.  This returns a new
global state in which the shared state and thread local states have
been updated to reflect the fact that all stores to memory and loads
from memory to thread-local storate have completed.

@@

\[
\inferrule{
w_i(j) = (l, \mathit{ss})
\\
(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l', \mathit{ss}'))
\\
w_i' = w_i[j \mapsto (l', \mathit{ss}')]
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma', (\dots, w_i', \dots))
}
(\textsc{Thread-Step})
\]

@@

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerkill();\mathit{ss})
\\
w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
}
(\textsc{Kill-No-Op})
\]

@@

\[
\inferrule{
\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
\\
M > 0
}
{
(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
}
(\textsc{Kill})
\]

@@

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
\\
w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
\\
k \in [0, N - M]
\\
\forall a \in [1, k] \;.\; w_{M+a} = ((l_1, \mathit{ss}), \dots, (l_1, \mathit{ss}))
}
{
(\sigma, (\dots, w_i, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_M, w_{M+1}, \dots, w_{M+k}, \bot, \dots, \bot))
}
(\textsc{Fork})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
\\
\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow \mathsf{sync}(\sigma , (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Barrier})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
\\
\forall j\;.\;w_1'(j) = (l_{1,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\mathit{ss})
\\
\forall i \neq 1 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Resizing-Barrier})
\]

@@

\caption{Abstract operational semantics for our cooperative kernels language extensions}

@@

>>>>

\newcommand{\myss}{\mathit{ss}}
\newcommand{\Stmts}{\mathsf{Stmts}}
\newcommand{\threadstates}{\mathsf{ThreadStates}}
\newcommand{\sharedstates}{\mathsf{SharedStates}}
\newcommand{\sync}{\mathsf{sync}}

In \mysec\ref{sec:semantics} we presented the semantics of cooperative
kernels relatively informally, using English, to provide the intuition
behind our programmming model.  We now back this up with a more formal
presentation as an operational semantics for an abstract GPU
programming model.

\paragraph{States}

Let $L$ be a set of \emph{local states} that abstractly captures the
private memory associated with a thread executing a GPU kernel.  Let
$\Stmts$ denote the set of all possible statements that a
thread can execute.  We do not detail the structure of these
statements, except that we assume sequential composition of statements
is provided by the $\code{;}$ separator, and that the $\offerkill$,
$\offerfork$, $\globalbarrier$ and $\resizingglobalbarrier$ primitives
from our cooperative kernels programming model are valid statements.

A \emph{thread state} is then a pair $(l, \myss)$, where $l \in L$ and
$\myss \in \Stmts$.  The $l$ component captures the valuation of all the
thread's private memory, and the $\myss$ component captures the
remaining statements to be executed by the thread.  Let $\threadstates$ denote the set of all thread states.

Assuming that $d > 0$ threads per workgroup were requested on kernel launch, a \emph{workgroup state}
is a $d$-tuple $((l_0, \myss_0), \dots, (l_{d-1}, \myss_{d-1}))$, where each $(l_i, \myss_i)$ is the thread state for the $i$th thread in the workgroup $(0\leq i < d$).

Assuming that $N > 0$ was specified as the maximum number of
workgroups that should execute the cooperative kernel, a \emph{kernel
  state} is then a pair
%
\[
(\sigma, (w_0, \dots, w_{M-1}, \underbrace{\bot, \dots,
\bot}_{N-M}))\]
%
where: $\sigma$ represents the shared state of the kernel; $M
\leq N$ is the number of \emph{active} workgroups; $w_i$ is the
workgroup state for active workgroup $i$ ($0 \leq i < M$); and $N-M$
occurrences of $\bot$ indicate \emph{absent} workgroups.  Let $\sharedstates$ denote the set of all possible shared states.

\paragraph{Thread-level transitions}

We leave the semantics for thread-level transitions abstract, assuming
a binary relation $\rightarrow_{tau}$ on $\sharedstates{} \times
\threadstates{}$.  If $(\sigma, (l, \myss)) \rightarrow_{\tau}
(\sigma', (l', \myss'))$, this indicates that if a thread is in local
state $(l, \myss)$, the thread can transition to local state $(l',
\myss')$, changing the shared state from $\sigma$ to $\sigma'$ in the
process.

All we require is that $(\sigma, (l, \myss)) \rightarrow_{\tau}
(\sigma', (l', \myss'))$ if $\myss$ has the form $\mathsf{special}();
\mathit{tt}$, where $\mathsf{special}$ is one of $\offerkill$,
$\offerfork$, $\globalbarrier$ or $\resizingglobalbarrier$.  This is
because we shall specifically define the meaning of the new primitives
introduced by our programming model.

\paragraph{Memory synchronisation}

GPUs are known to have relaxed memory models~\cite{ }.  To abstractly
account for this, we assume that the shared state component $\sigma$
is not simplyl a mapping from locations to values, but instead
captures all the intricacies of GPU memory that can lead to this
relaxed behaviour.  We also assume a function $\sync$ which, given a
kernel state $\kappa$, returns a set of kernel states.  The idea is
that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
can be reached from $\kappa$ by stalling until all stores to memory
and loads from memory to thread-local storage have completed.  All we
require is that $\sync$ does not modify the number of active
workgroups nor the component of a thread state that determines which
statements remain to be executed.

@@

\[
\inferrule{
w_i(j) = (l, \myss)
\\
(\sigma, (l, \myss)) \rightarrow_{\tau} (\sigma', (l', \myss'))
\\
w_i' = w_i[j \mapsto (l', \myss')]
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma', (\dots, w_i', \dots))
}
(\textsc{Thread-Step})
\]

@@

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerkill();\myss)
\\
w_i' = ((l_0, \myss), \dots, (l_{d-1}, \myss))
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
}
(\textsc{Kill-No-Op})
\]

@@

\[
\inferrule{
\forall j \;.\; w_{M-1}(j) = (l_j, \offerkill();\myss)
\\
M > 0
}
{
(\sigma, (\dots, w_{M-2}, w_{M-1}, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-2}, \bot, \bot, \dots, \bot))
}
(\textsc{Kill})
\]

@@

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerfork();\myss)
\\
w_i' = ((l_{0}, \myss), \dots, (l_{d-1}, \myss))
\\
k \in [0, N - M]
\\
\forall a \in [0, k - 1] \;.\; w_{M+a} = ((l_{0}, \myss), \dots, (l_{0}, \myss))
}
{
(\sigma, (\dots, w_i, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_{M-1}, w_{M}, \dots, w_{M+k-1}, \bot, \dots, \bot))
}
(\textsc{Fork})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\myss)
\\
\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \myss)
\\
\kappa \in \sync((\sigma , (w_{0}', \dots, w_{M-1}', \bot, \dots, \bot))
}
{
(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow \kappa)
}
(\textsc{Barrier})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\myss)
\\
\forall j\;.\;w_{0}'(j) = (l_{1,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\myss)
\\
\forall i \neq 1 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\myss)
}
{
(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow (\sigma, (w_{0}', \dots, w_{M-1}', \bot, \dots, \bot))
}
(\textsc{Resizing-Barrier})
\]

@@

\caption{Abstract operational semantics for our cooperative kernels language extensions}\label{fig:semanticrules}

@@

\paragraph{Operational semantics rules}

\myfiglong\ref{fig:semanticrules} presents the rules of our
operational semantics, defining a relation $\rightarrow$ on kernel
states.

Rule $\textsc{Thread-Step}$ defines the semantics for thread making a
single execution step, delegating to the abstract $\rightarrow_{\tau}$
relation to determine how the thread's local state and the shared
state component change.

Rule $\textsc{Kill-No-Op}$ reflects the fact that when a workgroup
reaches $\offerkill$, this can always be treated as a no-op.  Whether
a scheduler implementation accepts $\offerkill$ calls or not depends
on competing workloads and how the scheduler has been designed to meet
the non-functinoal requirements discussed in
\mysec\ref{sec:nonfunctional}, but in general the programmer should
always be prepared for the possiblity that a workgroup survives after
calling $\offerkill$.

The case where a workgroup's offer to be killed is accepted by the
scheduler is captured by rule $\textsc{Kill}$.  Because we have
adopted a semantics where workgroup 0 is never killed and where only
the workgroup with the highest id can be killed, the rule only fires
if $M > 0$ and workgroup $M-1$ has reached $\offerkill$.  The rule has
the effect of replacing the workgroup state for $w_{M-1}$
with $\bot$.

Recall that $\offerkill$ is a workgroup-level function: the same
syntactic $\offerfork$ call must be reached by all threads in a
workgroup.  This is captured in our rules by requiring in both
$\textsc{Kill-No-Op}$ and $\textsc{Kill}$ that every thread is ready
to execute $\offerkill$ followed by an identical statement $\myss$.
Neither rule applies until all threads in a workgroup reach
$\offerfork$, and the workgroup gets stuck if multiple threads in a
workgroiup reach $\offerfork$ with different subsequent statements.

\ADComment{TODO: $\textsc{Fork}$}

\ADComment{TODO: $\textsc{Barrier}$}

\ADComment{TODO: $\textsc{Resizing-Barrier}$}

\paragraph{Simplifications}

\ADComment{All state is transmitted.}  \ADComment{Local memory is not
  explicitly modeled but can be captured inside shared state.}
\ADComment{Semantics for intra-workgroup barriers are ignored.}
